Parameter Inference for a Simple Stochastic Hydrological Model

### 2015-02-18 TeXFiles/stochInf1D.tex
First upload of Tex files

The dynamics of a linear reservoir with state-dependent noise can be described by a stochastic differential equation, interpreted according to the Stratonovich convention \textbf{[ref?]},

\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}

where $r(t)$ describes the rain input, $S(t)$ is the volume of water in the system, $K$ is the {\em retention time} and the dimensionless parameter $\gamma$ quantifies the noise ${\eta}(t)$. Henceforth we shall assume standard white noise, i.e.,

\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}

The term $\left(1+\gamma/2\right)$ in (\ref{sde}) arises from the Stratonovich interpretation. According to \cite{lau_2007_StateDepDiff}, the probability $P(S)$ of finding the system in a state $S(T)$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as,

\begin{equation}\label{pathint}
P(S) = \int_{S_0}^{S(T)} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{DS} \,,
\end{equation}

where the integral extends over all paths starting at $S_0$ and ending at $S$. Following \cite{lau_2007_StateDepDiff}, one can derive from (\ref{sde}) and (\ref{whitenoise}) the formal path-measure,

\begin{equation}\label{pathmeasure}
{\mathcal DS}
=
\prod_{t}
\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
\end{equation}

defined as an infinite product in the formal limit $\Delta t \rightarrow 0$, and the {\em action} $\tilde{\mathcal S}:[0,T]\longrightarrow \mathbb R_+$,

\begin{equation}\label{action}
  \tilde{\mathcal S}[\dot{S(t)},S(t)]
  =
  \int_0^T
  dt
  \left\{
    \frac{K}{2\gamma S^2(t)}
    \left(
        \dot S(t)
        -
        r(t)
        +
        \frac{1+\gamma}{K}S(t)
    \right)^2
    -
    \frac{2+\gamma}{4K}
  \right\}
 \,.
\end{equation}

Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S) \mathit{dS}}=1$.

It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S(T))$ in the simple case of a constant rain input $r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,

\begin{equation}\label{detailed_balance}
P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,,
\end{equation}

where $P(S_1 t_1 | S_0 t_0 )$ is the conditional probability for the state $S_1$ at time $t_1$ given the initial state $S_0$ at time $t_0$ and $P(S_0 t_1 | S_1 t_0 )$ is the conditional probability for the reverse path $\overline{S} = S(-t)$. Assuming a solution in the form

@@

\begin{equation}\label{detailed_balance_2}
P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 ) \exp^{-\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{\partial \mathcal{H}}{\partial S}}\,.
\end{equation}

@@

After the transformation $S(t) = K r(t) e^{q(t)}$, the posterior reads

\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -{\tilde{\mathcal S}}[\dot{q(t)},q(t)]
    -\frac{1}{2}
    \sum_{i=1}^n
    \frac
    {(y_i-r_ie^{q_i})^2}
    {\sigma^2}
    -\ln(K\gamma)
  \bigg]
  {\mathcal Dq}({\pmb\theta})
  \,,
\end{equation}

%\prod_{t}
%\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,

where $r_i$ and $q_i$ represent $r(t)$ and $q(t)$, respectively, at time $t_i$. In the transformed variables the path-measure takes the form, up to a parameter-independent multiplicative factor,

\begin{equation}\label{pathmeasure_q}
{\mathcal Dq} (\bt)
=
\prod_{t}
\sqrt{\frac{K}{\gamma}} dq(t)\,,
\end{equation}

@@

\begin{equation}\label{action_new}
{\mathcal S}[\dot{q},q]
=
\frac{1}{2\gamma K}
\int_0^T \dt \left\{\left(
    K^2\dot q^2(t) +
    \rho(t)-e^{-q(t)}\right)^2 -
    \frac{\gamma(2+\gamma)}{2}
\right\} \,,
\end{equation}

with
\begin{equation}\label{rho}
\rho(t)=K\frac{\dot r(t)}{r(t)}+\gamma+1 = K\frac{d \left[\ln r(t)\right]}{dt}+\gamma+1\,.
\end{equation}

@@

\begin{equation}\label{H}
  \mathcal{H}_q(q,t)
  =
  e^{-q}+q\rho(t)\,,
\end{equation}

@@

\begin{multline}\label{action_final}
{\mathcal S}[\dot{q},q]
=
\frac{1}{2\gamma K}
\int_0^T dt\left\{
    K^2\dot q^2(t) +
    (\rho(t)-e^{-q(t)})^2 -
    2K\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
    \frac{\gamma(2+\gamma)}{2}
\right\}
\\
+
\frac{1}{\gamma}\left(
    \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
\right)
\\
= \frac{1}{2\gamma K}
\int_0^T dt\left\{
    K^2\dot q^2(t) +
    (\rho(t)-e^{-q(t)})^2 -
    2Kq(t)\dot\rho(t) -
    \frac{\gamma(2+\gamma)}{2}
\right\}
\\
+
\frac{1}{\gamma}\left(
    e^{-q(T)}+q(T)\rho(T)
    -e^{-q(0)}-q(0)\rho(0)
\right)
\,.
\end{multline}

>>>>

In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the {\em retention time}, and the noise.
The noise parameter is dimensionless, which renders the noise term scale-invariant.
Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
For computational reasons, we work with the {\em Stratonovich} convention.
Our parametrization is such that, for constant rain input $r(t)=r$, the mean of $S$ converges to $Kr$, in the long-time limit (see below).

According to \cite{lau_2007_StateDepDiff}, the probability $P(S)$ of finding the system in a state $S$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as,

\begin{equation}\label{pathint}
P(S) = \int_{S_0}^{S} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{D}S \,,
\end{equation}
where the integral extends over all paths starting at $S_0$ and ending at $S$.
The path-measure $\mathcal DS$ is the formal limit $\Delta t\rightarrow 0$ of the discretized measure
\begin{equation}\label{pathmeasure}
{\mathcal D}S
=
\prod_{i}
\frac{dS_i}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S_i}\,.
\end{equation}
The {\em action} is a functional defined on paths $\tilde{\mathcal S}:[0,T]\longrightarrow \mathbb R_+$ and reads as
\begin{equation}\label{action}
  \tilde{\mathcal S}[\dot{S}(t),S(t)]
  =
  \int_0^T
  dt
  \left\{
    \frac{K}{2\gamma S^2(t)}
    \left(
        \dot S(t)
        -
        r(t)
        +
        \frac{1+\gamma}{K}S(t)
    \right)^2
    -
    \frac{2+\gamma}{4K}
  \right\}
 \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S) \mathit{dS}}=1$.

It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S(T))$ in the simple case of a constant rain input $r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,
\begin{equation}\label{detailed_balance}
P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,,
\end{equation}
where $P(S_1 t_1 | S_0 t_0 )$ is the conditional probability for the state $S_1$ at time $t_1$ given the initial state $S_0$ at time $t_0$ and $P(S_0 t_1 | S_1 t_0 )$ is the conditional probability for the reverse path $\overline{S}(t) = S(-t)$. Assuming a solution in the form

@@

\begin{equation}\label{detailed_balance_2}
P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 ) \exp^{-\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{\mathcal{H}}{S}}\,.
\end{equation}

@@

We replace the variable $S(t)$ and the parameter $K$ by the dimensionless quantities, $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
In terms of these dimensionless variables and parameters the posterior reads (without prior)
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -{\tilde{\mathcal S}}[\dot{q}(t),q(t)]
    -\frac{1}{2}
    \sum_{i=1}^n
    \frac
    {(y_i-r_ie^{\beta q_i})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,,
\end{equation}
%\prod_{t}
%\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
where $r_i$ and $q_i$ represent $r(t)$ and $q(t)$, respectively, at time $t_i$. In the transformed variables the path-measure takes the form, up to a parameter-independent multiplicative factor,

\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,,
\end{equation}

@@

\begin{equation}\label{action_new}
\tilde {\mathcal S}[\dot{q},q]
=
\frac{1}{2T}
\int_0^T \dt \left\{\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{(2+\gamma)\beta^2}{2\gamma}
\right\} \,,
\end{equation}

with
\begin{equation}\label{rho}
\rho(t)=\frac{T\dot r(t)}{\beta r(t)}+\frac{\beta(\gamma+1)}{\gamma}
=
\frac{T}{\beta}\frac{d \left[\ln r(t)\right]}{dt}+\frac{\beta(\gamma+1)}{\gamma}\,.
\end{equation}

@@

\begin{equation}\label{H}
  \mathcal{H}_q(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}

@@

\begin{multline}\label{action_final}
{\mathcal S}[\dot{q},q]
=
\frac{1}{2T}
\int_0^T dt\left\{
    T^2\dot q^2(t) +
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    2T\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
     \frac{(2+\gamma)\beta^2}{2\gamma}
\right\}
\\
+ \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
\\
= \frac{1}{2T}
\int_0^T dt\left\{
    T^2\dot q^2(t) +
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    2Tq(t)\dot\rho(t) -
    \frac{(2+\gamma)\beta^2}{2\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}
